---
title: "Analysis on the relationship of Population against COVID-19 cases and Deaths
  in the US by County"
author: "Mitchell"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

# Summary
The analysis shows that the population of US counties have a linear relationship to the number of cases and deaths experienced by that county.

# About the dataset
The data found in this document comes from the repository for the 2019 Novel Coronavirus Visual Dashboard operated by the Johns Hopkins University Center for Systems Science and Engineering (JHU CSSE). The data extracted describes the total cases and deaths in the US, split by states & counties, updated daily.

# Data Preparation
First, we will start by importing some useful packages.

```{r import libraries}
library(tidyverse)
library(ggplot2)
```

Start by reading the data from the 2 CSV files, relating to the US cases.

```{r get_jhu_data}
## Get current data in 2 files, they start the same way
url_in <- 
  "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"
file_names<-
  c("time_series_covid19_confirmed_US.csv", "time_series_covid19_deaths_US.csv")
urls <- str_c(url_in, file_names)
```

Let's read in the data and see what we have.

```{r import_data, message = FALSE}
US_cases <- read_csv(urls[1])
US_deaths <- read_csv(urls[2])
```

Now, time to tidy up the datasets. I decided that I didn't need the iso codes, FIPS, code3, or Country_Region, as the analysis was all US based.
I was not intending to analyse Lat/Long data, so I removed that.
Finally, I removed all the dates except for the last available date (3/9/2023), to get the final number of COVID cases & COVID deaths before the dataset had stopped updating.

```{r tidy_US_data}
US_cases <- US_cases %>%
  select(c(UID,Admin2,Province_State,Combined_Key,'3/9/23'))

colnames(US_cases)[5] <- "cases"

US_deaths <- US_deaths %>%
  select(c(UID,Admin2,Province_State,Combined_Key,Population,'3/9/23'))

colnames(US_deaths)[6] <- "deaths"

US_county_data <- left_join(US_cases,US_deaths,by = c("UID", "Admin2",
                                                      "Province_State","Combined_Key"))
```

# Data Summary
Having grouped all of the county data together, let us see a quick summary of the data at hand.

```{r summary US data}
summary(US_county_data)
```

# Top 10 counties in cases and deaths
Now, we will take a look at the top 10 counties in terms of cases & deaths and their relative population sizes.
```{r Top 10 counties for cases and deaths}
US_county_data <- US_county_data %>%
  mutate(population_rank = dense_rank(desc(Population)))

US_county_data %>%
  top_n(10, cases) %>% # Select top 10 counties by cases
  ggplot(aes(x = reorder(Combined_Key, cases), y = cases, fill = Province_State)) +
  geom_col() +
  geom_text(aes(label = paste0("Rank: ", population_rank)), # Add rank as labels
            hjust = -0.2, # Position text slightly outside the bars
            size = 3) +   # Adjust text size
  coord_flip() + # Flip coordinates for better readability
  labs(
    title = "Top 10 Counties by COVID-19 Cases",
    x = "County",
    y = "Number of Cases",
    fill = "State"
  ) +
  theme_minimal()

US_county_data %>%
  top_n(10, cases) %>% # Select top 10 counties by cases
  ggplot(aes(x = reorder(Combined_Key, cases), y = cases, fill = Province_State)) +
  geom_col() +
  geom_text(aes(label = paste0("Rank: ", population_rank)), # Add rank as labels
            hjust = -0.2, # Position text slightly outside the bars
            size = 3) +   # Adjust text size
  coord_flip() + # Flip coordinates for better readability
  labs(
    title = "Top 10 Counties by COVID-19 Deaths",
    x = "County",
    y = "Number of Deaths",
    fill = "State"
  ) +
  theme_minimal()

```

At the top end, it looks very much like the counties with the highest population have the most cases and deaths.

# Scatter plot of all counties
To help us better understand everything, perhaps a scatterplot visualisation of all counties may be helpful.

```{r scatterplot of all the counties}

ggplot(data = US_county_data, aes(x = Population,y = cases)) +
  geom_point(alpha = 0.7) +
  scale_size(range = c(2, 10)) +
  labs(
    title = "COVID-19 Cases and Population by County",
    x = "Population",
    y = "Number of Cases",
  ) +
  theme_minimal()

ggplot(data = US_county_data, aes(x = Population,y = deaths)) +
  geom_point(alpha = 0.7) +
  scale_size(range = c(2, 10)) +
  labs(
    title = "COVID-19 Deaths and Population by County",
    x = "Population",
    y = "Number of Deaths",
  ) +
  theme_minimal()

```

It does look like there is a strong linear relationship between a county's population and its case volume.
There also seems to be a linear relationship between a county's population and it's death toll, though visually it does seem less strong.

# Linear Modelling of Population against cases & deaths
Let us see how appropriate a linear model is for explaining the relationship between population and cases or deaths.

```{r linear modelling analysis for cases}

# Linear model: Population vs Cases
lm_cases <- lm(cases ~ Population, data = US_county_data)

# Summary of the linear model for cases
summary(lm_cases)

# Scatter plot for cases
ggplot(US_county_data, aes(x = Population, y = cases)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  labs(title = "Population vs Cases",
       x = "Population",
       y = "Number of Cases") +
  theme_minimal()

qqnorm(residuals(lm_cases))      # Generate the Q-Q plot
qqline(residuals(lm_cases), col = "blue")  # Add a reference line

```

```{r linear modelling analysis for deaths}
# Linear model: Population vs Deaths
lm_deaths <- lm(deaths ~ Population, data = US_county_data)

# Summary of the linear model for deaths
summary(lm_deaths)

# Scatter plot for deaths
ggplot(US_county_data, aes(x = Population, y = deaths)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE, color = "red") +
  labs(title = "Population vs Deaths",
       x = "Population",
       y = "Number of Deaths") +
  theme_minimal()

qqnorm(residuals(lm_deaths))      # Generate the Q-Q plot
qqline(residuals(lm_deaths), col = "red")  # Add a reference line

```

Based on the extremely high R^2 and extremely small p value, population is an extremely strong predictor of cases and deaths.
Furthermore, the factors are likely to be linearly related, though review of the q-q plot shows that the underlying data has more extreme values than expected, implying significant violations of normality in the base dataset itself.

# Conclusion
Based on the analysis performed above, the relationship between populations and cases/deaths is likely linear. However, the normality violations indicated by the Q-Q plot suggest that the model assumptions might be compromised, especially for inference.

# Biases
My primary bias is that I expected some sort of relationship between population and cases/deaths, and attempted to mitigate it by going a step further to review the q-q plot of the data after finding such a good fit in the linear model. To my surprise there seems to be some violations, which may imply a different relationship rather than just a linear one.

```{r session info}
sessionInfo()
```

